---
title: 'MetaboHub project. Step0 Conversion matlab code into R code. 
Exploration functions and code of the Consensus OPLS-DA method.'
author: "Celine Bougel"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: kable
    highlight: pygments
    number_sections: yes
    self_contained: yes
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
  editor_options:
    chunk_output_type: console
  pdf_document:
    toc: yes
    toc_depth: '3'
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
#install.packages("knitr")
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
set.seed(12)
```

Before any action, it is necessary to verify that the needed packages are 
installed (the code chunks are not shown) and to define define the localisation 
of the codes and the files.

```{r packages_installation, class.source='fold-hide', warning=FALSE, include=FALSE, message=FALSE}
#install.packages(c("R.matlab", "DiagrammeR", "kopls"))
#update.packages(ask = FALSE)
```

```{r packages_load, class.source='fold-hide', warning=FALSE, include=FALSE, message=FALSE}
library(R.matlab) # to read MATLAB data
library(DiagrammeR) # to make mermaid diagrams
```

```{r access}
path_data <- "../../../data/data_brut" 
path_results <- "../../../results" 
path_functions <- "../functions"
```

```{r theme_ggplot2, class.source='fold-hide'}
# theme_graphs <- theme_bw() + theme(strip.text = element_text(size=14),
#                                    axis.title = element_text(size=16),
#                                    axis.text = element_text(size=14),
#                                    plot.title = element_text(size = 16),
#                                    legend.title = element_text(size = 14))
```


# Content of this file: Traduction of matlab code

In this file, the aim is to translate the `MATLAB` functions used in Julien 
Boccard's OPLS-DA consensus method (DOI: `10.1016/j.aca.2013.01.022`).



In this code example, the aim is to perform a discriminant analysis of NCI-60
cancer cell lines from two origins: colon versus ovary.
The data is divided into 3 blocks, evaluated on the same observations:

- Metabolomics: 14x 150 variables
- Microarray: 14x 200 variables
- Proteomics: 14x 100 variables





# Mind map of functions used

To see the function tree and how they relate to each other, see the README for 
the project's MIA forge.





# Loading functions

## Function `matrix2saisir`

```{r matrix2saisir}
source(file = file.path(path_functions, "Step0_1_matrix2saisir.R"))
```

## Function called by `ConcensusOPLSCV`

```{r kopls_functions_for_ConsensusOPLSCV}
#source(file = file.path(path_functions, "Step0_2_ConsensusOPLSCV.R"))
#source(file = file.path(path_functions, "Step0_2_1_koplsDummy.R"))
source(file = file.path(path_functions, "Step0_2_2_koplsReDummy.R"))
source(file = file.path(path_functions, "Step0_2_3_koplsCrossValSet.R"))
source(file = file.path(path_functions, "Step0_2_4_koplsScale.R"))
source(file = file.path(path_functions, "Step0_2_5_koplsCenterKTeTe.R"))
source(file = file.path(path_functions, "Step0_2_6_koplsCenterKTeTr.R"))
source(file = file.path(path_functions, "Step0_2_7_koplsCenterKTrTr.R"))
source(file = file.path(path_functions, "Step0_2_8_koplsModel.R"))
source(file = file.path(path_functions, "Step0_2_9_koplsPredict.R"))
source(file = file.path(path_functions, "Step0_2_10_koplsReScale.R"))
#source(file = file.path(path_functions, "Step0_2_11_koplsMaxClassify.R"))
#source(file = file.path(path_functions, "Step0_2_12_koplsBasicClassify.R"))
#source(file = file.path(path_functions, "Step0_2_13_koplsConfusionMatrix.R"))
#source(file = file.path(path_functions, "Step0_2_14_koplsSensSpec.R"))
source(file = file.path(path_functions, "Step0_2_15_koplsScaleApply.R"))
```

## Function called by `RVConcensusOPLS`

This function also uses `koplsModel` and `koplsScale`, which have already been loaded to run `ConcensusOPLSCV`.

```{r kopls_functions_for_RVConsensusOPLS}
#source(file = file.path(path_functions, "Step0_3_RVConsensusOPLS.R"))
source(file = file.path(path_functions, "Step0_3_1_koplsKernel.R"))
source(file = file.path(path_functions, "Step0_3_2_RVmodified.R"))
#source(file = file.path(path_functions, "Step0_3_3_DQ2.R"))
```

## Function called by `RVConcensusOPLSPerm`

This function also uses `RVmodified`, which have already been loaded to run 
`RVConsensusOPLS`. 

```{r kopls_functions}
# source(file = file.path(path_functions, "Step0_4_RVConcensusOPLSPerm.R"))
```





# Data preprocessing on `Demo_data` file

```{r import_demo_data.mat}
demo_data <- R.matlab::readMat(con = file.path(path_data, "demo_data.mat"))
```

This file corresponds to a `r class(demo_data)` of size `r length(demo_data)`.
It contains the data of `r ls(demo_data)`.

## Rename Block names

```{r names_data}
demo_data$BlockNames <- c("MetaboData", "MicroData", "ProteoData")
```

## Unit variance scaling

To use the OPLS-DA consensus method, it is necessary to calculate the Z-score of 
the data, i.e. each columns of the data are centered to have mean 0 and scaled
to have standard deviation 1.

```{r created_collection_data}
# Extract the 3 blocks of data
collection <- lapply(X = 1:length(demo_data$BlockNames), 
                     FUN = function(X){
                       # Extract the tables of data
                       name_data <- demo_data$BlockNames[X]
                       
                       # Check identical values for all subjects and delete
                       col_identic_val <- base::apply(X = demo_data[[name_data]],
                                                      MARGIN = 2, 
                                                      FUN = function(X){
                                                        base::length(base::unique(X)) < 2})
                       if(length(which(col_identic_val)) != 0){
                         print(paste0("The columns ", paste(which(col_identic_val),
                                                            sep = " ", 
                                                            collapse = ", "), 
                                      " of the ", name_data, 
                                      " data contained identical values."))
                         data <- demo_data[[name_data]][, -which(col_identic_val)]
                       } else{
                         print(paste0("The ", name_data, 
                                      " data does not contain columns with identical values"))
                         data <- demo_data[[name_data]]
                       }
                       
                       # Calculate the Z-score 
                       # (mean centering and standard deviations scaling)
                       scaled_data <- base::scale(data, center = TRUE,
                                                  scale = TRUE)
                       # Apply the function matrix2saisir
                       matrix <- matrix2saisir(data = scaled_data,
                                               coderow = NA, codecol = NA,
                                               position_coderow = NA,
                                               position_codecol = NA)
                       return(matrix)
                       })

# Rename blocks with BlokNames
names(collection) <- demo_data$BlockNames
ls(collection)
```

The results were observed (results not shown).

```{r check_collection_objects, class.source='fold-hide'}
# View(collection[[names(collection)[1]]])
# View(collection[[names(collection)[2]]])
# View(collection[[names(collection)[3]]])
```

Missing data were observed in relation to the Matlab data. This is due to columns 
having strictly identical values for all individuals, resulting in an error when 
using the `scale` function. A check has therefore been added against the source 
code demonstrating the use of the method. The dimensions of the data after 
verification were therefore:

```{r new_dim_collection}
lapply(X = 1:length(collection),
       FUN = function(X){dim(collection[[X]])})
```

Another check: all data blocks must be the same size. So, implement a code stop
if this is not the case:

```{r verif_size}
size <- lapply(X = 1:length(collection),
               FUN = function(X){
                 dim(collection[[X]])[1]
               })
if (all(size == size[[1]])) {
  size <- size[[1]]
  paste0("All data sets have the same number of subjects: ", size)
} else {
  stop("Data sets do not have the same number of subjects: ", size)
}
```

The next step in the source code is to extract the variable names (columns from 
the previous table) according to the data table concerned.
As the dimensions differ from those in Matlab following the previous 
modifications, a new list is created to contain this information:

```{r extract_variables_names}
# Extraction of names matches for each data block
list_var <- names(demo_data)[grep("VarNames", names(demo_data))]

var_names <- lapply(X = 1:length(list_var), 
                    FUN = function(X){
                      name_var <- list_var[X]
                      extract <- as.data.frame(demo_data[[name_var]])
                      
                      # Modifies input format
                      apply(extract[1, ], MARGIN = 2,
                            FUN = function(X){
                              extract[1, ] <- as.character(unlist(X))
                            })
                      return(extract[1, ])
                    })

# Rename blocks with BlokNames
names(var_names) <- list_var
ls(var_names)
length(var_names[[1]])
length(var_names[[2]])
length(var_names[[3]])
```





# Consensus OPLS model

Computation of a consensusOPLS model with leave-one-out cross-validation.

```{r define_cv_parameters}
# Number of predictive component(s)
LVsPred <- 1

# Maximum number of orthogonal components
LVsOrtho <- 1

# Number of cross-validation folds
CVfolds <- size
CVfolds
```

```{r run_consensusOPLSmodel}
data = collection ; Y = demo_data$Y ; A = LVsPred
maxOrtholvs = LVsOrtho ; nrcv = CVfolds ; cvType = "nfold"
modelType = "da" ; verbose = TRUE

# RVConsensusOPLSModelCV <- RVConsensusOPLS(data = collection,
#                                           Y = demo_data$Y,
#                                           A = LVsPred,
#                                           maxOrtholvs = LVsOrtho,
#                                           nrcv = CVfolds,
#                                           cvType = "nfold",
#                                           modelType = "da",
#                                           verbose = TRUE)
```



## Define parameters

To start the method, we need to define some parameters:

```{r init_parameters}

```


## Main function

```{r}
# translation of RVConsensusOPLS function
```





# Plot the results

```{r}

```





# Permutations

```{r}

```
















# Reproducibility

```{r reproducibility}
sessionInfo()
```
