---
title: 'MetaboHub project. Step0 Conversion matlab code into R code. 
Exploration functions and code of the Consensus OPLS-DA method.'
author: "Celine Bougel"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: show
    df_print: kable
    highlight: pygments
    number_sections: yes
    self_contained: yes
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
  editor_options:
    chunk_output_type: console
  pdf_document:
    toc: yes
    toc_depth: '3'
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
#install.packages("knitr")
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
set.seed(12)
```

Before any action, it is necessary to verify that the needed packages are 
installed (the code chunks are not shown) and to define define the localisation 
of the codes and the files.

```{r packages_installation, class.source='fold-hide', warning=FALSE, include=FALSE, message=FALSE}
#install.packages(c("R.matlab", "DiagrammeR"))
#update.packages(ask = FALSE)
```

```{r packages_load, class.source='fold-hide', warning=FALSE, include=FALSE, message=FALSE}
library(R.matlab) # to read MATLAB data
library(DiagrammeR) # to make mermaid diagrams
```

```{r access}
path_data <- "../../../data/data_brut" 
path_results <- "../../../results" 
path_functions <- "../functions"
```

```{r theme_ggplot2, class.source='fold-hide'}
# theme_graphs <- theme_bw() + theme(strip.text = element_text(size=14),
#                                    axis.title = element_text(size=16),
#                                    axis.text = element_text(size=14),
#                                    plot.title = element_text(size = 16),
#                                    legend.title = element_text(size = 14))
```


# Content of this file: Traduction of matlab code

In this file, the aim is to translate the `MATLAB` functions used in Julien 
Boccard's OPLS-DA consensus method (DOI: `10.1016/j.aca.2013.01.022`).

In this code example, the aim is to perform a discriminant analysis of NCI-60
cancer cell lines from two origins: colon versus ovary.
The data is divided into 3 blocks, evaluated on the same observations:

- Metabolomics: 14x 150 variables
- Microarray: 14x 200 variables
- Proteomics: 14x 100 variables





# Mind map of functions used

```{r mermaid_mindMap, echo=FALSE, class.source='fold-hide'}
# mindmap
#     root((Consensus OPLS-DA <br/> method))
#         (matrix2saisir)
#             addcode
#         {{RVConsensusOPLS}}
#             1 koplsScale
#             2 koplsKernel
#             3 koplsModel
#             4 RV_modified
#             5 ConsensusOPLSCV
#             (6 DQ2)
#         {{RVConsensusOPLSPerm}}
#             1 RVConsensusOPLS
#             2 RV_modified
#         {{ConsensusOPLSCV}}
#             koplsDummy/ koplsReDummy
#             koplsCrossValSet
#             koplsScale/ koplsScaleApply
#             koplsCenterKT*
#             koplsModel
#             koplsPredict
#             koplsRescale
#             koplsMaxClassify/ koplsBasicClassify
#             koplsSensSpec
#             koplsConfusionMatrix
# ", height = '100%', width = '100%')
```





# Data preprocessing

## Function `matrix2saisir`

```{r matrix2saisir}
source(file = file.path(path_functions, "Step0_1_matrix2saisir.R"))
```

## `Demo_data` file

```{r import_demo_data.mat}
demo_data <- R.matlab::readMat(con = file.path(path_data,
                                               "demo_data.mat"))
```

This file corresponds to a `r class(demo_data)` of size `r length(demo_data)`.
It contains the data of `r ls(demo_data)`.

### Rename Block names

```{r names_data}
demo_data$BlockNames <- c("MetaboData", "MicroData", "ProteoData")
```

### Unit variance scaling

To use the OPLS-DA consensus method, it is necessary to normalize the data.

```{r created_collection_data}
# Extract the 3 blocks of data
collection <- lapply(X = 1:length(demo_data$BlockNames), 
                     FUN = function(X){
                       name_data <- demo_data$BlockNames[X]
                       scaled_data <- as.data.frame(scale(demo_data[[name_data]],
                                                          center = TRUE,
                                                          scale = TRUE))
                       matrix <- matrix2saisir(as.matrix(scaled_data))
                       return(matrix)
                     })

# Rename blocks with BlokNames
names(collection) <- demo_data$BlockNames
ls(collection)
```

```{r verif_size}
size <- lapply(X = 1:length(collection),
               FUN = function(X){
                 dim(collection[[X]])[1]
               })
size <- size[[1]]
```

```{r not_understand}
# je ne sais pas à quoi correspond ça :
# collection(1).i=MetaboVarNames;
# collection(2).i=MicroVarNames;
# collection(3).i=ProteoVarNames;
# ni quand c'est utilisé..
```





# Consensus OPLS model

Computation of a consensusOPLS model with leave-one-out cross-validation.

## Define parameters

To start the method, we need to define some parameters:

```{r init_parameters}
# Number of predictive component(s)
LVsPred <- 1

# Maximum number of orthogonal components
LVsOrtho <- 1

# Number of cross-validation folds
CVfolds <- size
```


## Main function

```{r}
# traduction fonction RVConsensusOPLS
```





# Plot the results

```{r}

```





# Permutations

```{r}

```
















# Reproducibility

```{r reproducibility}
sessionInfo()
```
