---
title: 'A consensus orthogonal partial least squares discriminant analysis
(OPLS-DA) strategy for multiblock Omics data fusion.'
author: "Celine Bougel"
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography: 20230901934_MTH2.0WP1.4_step1_1_vignette.bib
biblio-style: apalike
link-citations: yes
nocite: '@*'
output:
  html_document:
    code_folding: show
    df_print: kable
    highlight: pygments
    number_sections: yes
    self_contained: yes
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
  editor_options:
    chunk_output_type: console
---

<style>
body {
text-align: justify}
</style>

# The Consensus OPLS method

The `ConsensusOPLS` R package implements the translated `Matlab` version of the 
**Consensus OPLS(-DA)** approach [@BOCCARD2013] (available at [Gitlab repository](https://gitlab.unige.ch/Julien.Boccard/consensusopls)), with the 
kernel-based reformulations of the **NIPALS** algorithm [@LINDGREN1993]. It 
includes updated codes from the `KOPLS` R package: codes translated from the 
`Matlab` version  were then compared and corrected, if necessary, according to 
the package source codes [@BYLESJO2008] (available at [Github repository](https://github.com/sdechaumet/ramopls/tree/master/inst/package)).

<span style="color:blue"> Fortement inspiré de la vignette ropls!</span> The 
current package includes some quality metrics for the permutation models, as 
**R2**, **Q2**, **DQ2** [@WESTERHUIS2008], <span style="color:blue"> the 
permutation **diagnostics** (Szymanska et al. 2012), the computation of the 
**VIP values** (Wold, Sjostrom, and Eriksson 2001), the score and orthogonal 
distances to detect outliers (Hubert, Rousseeuw, and Vanden Branden 2005), as 
well as many graphics (scores, loadings, predictions, diagnostics, outliers, 
etc).</span>





# Context

Once the functions have been implemented on R, the aim was to perform a 
discriminant analysis on the `demo_data` set proposed by Julien Boccard in his 
method. This analysis concerned a selection of data from a publicly available 
repository of the National Cancer Institute, i.e. the NCI-60 dataset. This data 
set provides experimental data obtained from 60 human cancer cell lines derived 
from nine tissue origins [@SHOEMAKER2006]. Two tissue were selected in this 
study (colon versus ovary), and three data sources were chosen: transcriptomics,
proteomics and metabolomics data (3 blocks of data, evaluated on the same 
observations):

- Metabolomics: 14x 150 variables
- Microarray: 14x 200 variables
- Proteomics: 14x 100 variables

The combinaison of these data sources was excepted to provide a global profiling 
of the cell lines in an integrative systems biology perspective. The Consensus 
OPLS-DA strategy was applied for the differential analysis of the two selected 
tumour origins and the simultaneous analysis of the three blocks of data.





# R environment preparation

```{r setup, class.source='fold-hide'}
#install.packages("knitr")
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

# To ensure repeatability
set.seed(12)
```

Before any action, it is necessary to verify that the needed packages are 
installed (the code chunks are not shown) and to define define the localisation
of the codes and the files.

```{r access}
path_data <- "../../../data/data_brut" 
path_results <- "../../../results" 
path_functions <- "../functions"
```

The code below has been designed to have as few dependencies as possible on R 
packages, except for the stable packages.

```{r packages_installation, class.source='fold-hide', warning=FALSE, include=FALSE, message=FALSE}
#install.packages(c("R.matlab", "ggplot2", "ggrepel", "stats", "utils"))
#update.packages(ask = FALSE)
```

```{r packages_load, warning=FALSE, message=FALSE, class.source='fold-hide'}
library(R.matlab) # to read MATLAB data
library(ggplot2) # to make beautiful graphs
library(ggrepel) # to annotate ggplot2 graph
library(stats) # to use R statistical functions
library(utils) # to use R utility functions
```

Then we create a uniform theme that will be used for all graphic output.

```{r theme_ggplot2, class.source='fold-hide'}
theme_graphs <- theme_bw() + theme(strip.text = element_text(size=14),
                                   axis.title = element_text(size=16),
                                   axis.text = element_text(size=14),
                                   plot.title = element_text(size = 16),
                                   legend.title = element_text(size = 14))
```





# Loading functions

To execute the ConsensusOPLS method, the user must load all its functions.

**Function `matrix2saisir`**

```{r matrix2saisir, class.source='fold-hide'}
source(file = file.path(path_functions, "Step0_1_matrix2saisir.R"))
```

**Functions called by `ConcensusOPLSCV`**

```{r kopls_functions_for_ConsensusOPLSCV, class.source='fold-hide'}
source(file = file.path(path_functions, "Step0_2_ConsensusOPLSCV.R"))
source(file = file.path(path_functions, "Step0_2_1_koplsDummy.R"))
source(file = file.path(path_functions, "Step0_2_2_koplsReDummy.R"))
source(file = file.path(path_functions, "Step0_2_3_koplsCrossValSet.R"))
source(file = file.path(path_functions, "Step0_2_4_koplsScale.R"))
source(file = file.path(path_functions, "Step0_2_5_koplsCenterKTeTe.R"))
source(file = file.path(path_functions, "Step0_2_6_koplsCenterKTeTr.R"))
source(file = file.path(path_functions, "Step0_2_7_koplsCenterKTrTr.R"))
source(file = file.path(path_functions, "Step0_2_8_koplsModel.R"))
source(file = file.path(path_functions, "Step0_2_9_koplsPredict.R"))
source(file = file.path(path_functions, "Step0_2_10_koplsReScale.R"))
source(file = file.path(path_functions, "Step0_2_11_koplsMaxClassify.R"))
source(file = file.path(path_functions, "Step0_2_12_koplsBasicClassify.R"))
source(file = file.path(path_functions, "Step0_2_13_koplsConfusionMatrix.R"))
source(file = file.path(path_functions, "Step0_2_14_koplsSensSpec.R"))
source(file = file.path(path_functions, "Step0_2_15_koplsScaleApply.R"))
```

**Functions called by `RVConcensusOPLS`**

This function also uses `koplsModel` and `koplsScale`, which have already been 
loaded to run `ConcensusOPLSCV`.

```{r kopls_functions_for_RVConsensusOPLS, class.source='fold-hide'}
source(file = file.path(path_functions, "Step0_3_RVConsensusOPLS.R"))
source(file = file.path(path_functions, "Step0_3_1_koplsKernel.R"))
source(file = file.path(path_functions, "Step0_3_2_RVmodified.R"))
source(file = file.path(path_functions, "Step0_3_3_DQ2.R"))
```

**Function for VIP**
```{r VIP_function, class.source='fold-hide'}
source(file = file.path(path_functions, "Step0_5_VariableImportanceProjection.R"))
```


**Function called by `RVConcensusOPLSPerm`**

This function also uses `RVmodified`, which have already been loaded to run 
`RVConsensusOPLS`. 

```{r permutation_function, class.source='fold-hide'}
source(file = file.path(path_functions, "Step0_4_RVConsensusOPLSPerm.R"))
```





# Data preprocessing on `Demo_data` file

As mentioned earlier, we use the demonstration dataset proposed by Julien 
Boccard:

```{r import_demo_data.mat}
demo_data <- R.matlab::readMat(con = file.path(path_data, "demo_data.mat"))
```

This file corresponds to a `r class(demo_data)` of size `r length(demo_data)`.
It contains the data of `r ls(demo_data)`.

## Rename Block names

```{r names_data}
demo_data$BlockNames <- c("MetaboData", "MicroData", "ProteoData")
```

## Unit variance scaling

To use the OPLS-DA consensus method, it is necessary to calculate the Z-score of 
the data, i.e. each columns of the data are centered to have mean 0 and scaled
to have standard deviation 1.

```{r created_collection_data}
# Extract the 3 blocks of data
collection <- lapply(X = 1:length(demo_data$BlockNames), 
                     FUN = function(X){
                       # Extract the tables of data
                       name_data <- demo_data$BlockNames[X]
                       
                       # Extract columns length for all subjects
                       col_identic_val <- base::apply(X = demo_data[[name_data]],
                                                      MARGIN = 2, 
                                                      FUN = function(X){
                                                        base::length(base::unique(X)) < 2})
                       # Check identical values for all subjects and delete
                       if(length(which(col_identic_val)) != 0){
                         print(paste0("The columns ", paste(which(col_identic_val),
                                                            sep = " ", 
                                                            collapse = ", "), 
                                      " of the ", name_data, 
                                      " data contained columns with identical values."))
                         data <- demo_data[[name_data]][, -which(col_identic_val)]
                       } else{
                         print(paste0("The ", name_data, 
                                      " data doesn't contain columns with identical values."))
                         data <- demo_data[[name_data]]
                       }
                       
                       # Calculate the Z-score 
                       # (mean centering and standard deviations scaling)
                       scaled_data <- base::scale(data, center = TRUE,
                                                  scale = TRUE)
                       # Apply the function matrix2saisir
                       matrix <- matrix2saisir(data = scaled_data,
                                               coderow = NA, codecol = NA,
                                               position_coderow = NA,
                                               position_codecol = NA)
                       
                       # Return the result
                       return(matrix)
                       })
```

```{r rename_blocks}
# Rename blocks with BlokNames
names(collection) <- demo_data$BlockNames
ls(collection)
```

The results were observed (results not shown).

```{r check_collection_objects, class.source='fold-hide'}
# View(collection[[names(collection)[1]]])
# View(collection[[names(collection)[2]]])
# View(collection[[names(collection)[3]]])
```

Missing data were observed in relation to the `Matlab` data. This is due to 
columns having strictly identical values for all individuals, resulting in an 
error when using the `scale` function. A check has therefore been added against 
the source code demonstrating the use of the method. The dimensions of the data 
after verification were therefore:

```{r new_dim_collection}
lapply(X = 1:length(collection),
       FUN = function(X){dim(collection[[X]])})
```

Another check: all data blocks must be the same size. The user can see it in the 
output above, but let's set up a code stop if this is not the case:

```{r verif_size}
size <- lapply(X = 1:length(collection),
               FUN = function(X){
                 dim(collection[[X]])[1]
               })
if (all(size == size[[1]])) {
  size <- size[[1]]
  paste0("All data sets have the same number of subjects: ", size, ".")
} else {
  stop("Data sets do not have the same number of subjects: ", size, ".")
}
```

The next step in the source code is to extract the variable names (columns from 
the previous table) according to the data table concerned.
As the dimensions differ from those in Matlab following the previous 
modifications, a new list is created to contain this information:

```{r extract_variables_names}
# Extraction of names matches for each data block
list_var <- names(demo_data)[grep("VarNames", names(demo_data))]

var_names <- lapply(X = 1:length(list_var), 
                    FUN = function(X){
                      name_var <- list_var[X]
                      extract <- as.data.frame(demo_data[[name_var]])
                      
                      # Modifies input format
                      base::apply(X = extract[1, ], MARGIN = 2,
                                  FUN = function(X){
                                    extract[1, ] <- as.character(unlist(X))
                                  })
                      return(extract[1, ])
                    })

# Rename blocks with BlokNames
names(var_names) <- list_var
ls(var_names)
length(var_names[[1]])
length(var_names[[2]])
length(var_names[[3]])
```





# Consensus OPLS model

Computation of a consensusOPLS model with leave-one-out cross-validation.

```{r define_cv_parameters}
# Number of predictive component(s)
LVsPred <- 1

# Maximum number of orthogonal components
LVsOrtho <- 1

# Number of cross-validation folds
CVfolds <- size
CVfolds
```

```{r run_consensusOPLSmodel}
RVConsensusOPLSModelCV <- RVConsensusOPLS(data = collection,
                                          Y = demo_data$Y,
                                          A = LVsPred,
                                          maxOrtholvs = LVsOrtho,
                                          nrcv = CVfolds,
                                          cvType = "nfold",
                                          modelType = "da",
                                          verbose = TRUE)
```

The execution time for this part of the method takes `r RVConsensusOPLSModelCV$execution_time`.




# Display the main results

```{r print_main_results}
position <- RVConsensusOPLSModelCV$model$cv$OrthoLVsOptimalNum+1
paste0('R2: ', round(RVConsensusOPLSModelCV$model$Model$R2Yhat[position], 4))
paste0('Q2: ', round(RVConsensusOPLSModelCV$model$cv$Q2Yhat[position], 4))
paste0('DQ2: ', round(RVConsensusOPLSModelCV$model$cv$DQ2Yhat[position], 4))

#Confusion matrix
RVConsensusOPLSModelCV$model$da$confusMatrix
```

```{r print_VIP, class.source='fold-hide'}
VIP <- MBVIP(collection = collection, Y = demo_data$Y, 
             model = RVConsensusOPLSModelCV)
names(VIP) <- demo_data$BlockNames
```



# Plot the results

## Consensus Score plot

```{r ggplot_score_data, class.source='fold-hide', warning=FALSE}
ggplot2::ggplot(data = data.frame("x" = RVConsensusOPLSModelCV$model$Model$T,
                                  "y" = RVConsensusOPLSModelCV$model$Model$To,
                                  "Labels" = as.matrix(base::unlist(demo_data$ObsNames[, 1]))),
                aes(x = x, y = y, label = Labels, 
                    shape = Labels, colour = Labels)) +
  ggplot2::xlim(-0.5, 0.5)+
  ggplot2::ylim(-1, 1)+
  ggplot2::xlab("Predictive component") +
  ggplot2::ylab("Orthogonal component") +
  ggplot2::ggtitle("ConsensusOPLS Score plot")+
  ggplot2::geom_point(size = 2.5) + 
  ggrepel::geom_text_repel(size = 4, show.legend = FALSE) + 
  theme_graphs+
  ggplot2::scale_color_manual(values = c("#7F3C8D", "#11A579"))
```

## Block contributions to the predictive component

```{r ggplot_data_pred_compo, class.source='fold-hide'}
ggplot2::ggplot(data = data.frame("Values" = RVConsensusOPLSModelCV$model$Model$lambda[,1],
                                  "Blocks" = as.factor(labels(collection))),
                aes(x = Blocks, y = Values,
                    fill = Blocks, labels = Values)) +
  ggplot2::geom_bar(stat = 'identity') + 
  ggplot2::xlab("Data blocks") +
  ggplot2::ylab("Weight") +
  ggplot2::ggtitle("Block contributions to the predictive component")+
  theme_graphs+
  ggplot2::scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3"))
```

## Block contributions to the first orthogonal component

```{r ggplot_data_1st_ortho_compo, class.source='fold-hide'}
ggplot2::ggplot(data = data.frame("Values" = RVConsensusOPLSModelCV$model$Model$lambda[,2],
                                  "Blocks" = as.factor(labels(collection))),
                aes(x = Blocks, y = Values,
                    fill = Blocks, labels = Values)) +
  ggplot2::geom_bar(stat = 'identity') + 
  ggplot2::xlab("Data blocks") +
  ggplot2::ylab("Weight") +
  ggplot2::ggtitle("Block contributions to the first orthogonal component")+
  theme_graphs+
  ggplot2::scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3"))
```

## Block contributions: the two previous plots into one

```{r ggplot_data_pred_ortho, class.source='fold-hide'}
data_two_plots <- data.frame("Values" = RVConsensusOPLSModelCV$model$Model$lambda[,1],
                             "Type" = "Pred",
                             "Blocks" = labels(collection))
data_two_plots <- data.frame("Values" = c(data_two_plots$Values,
                                          RVConsensusOPLSModelCV$model$Model$lambda[,2]),
                             "Type" = c(data_two_plots$Type,
                                        rep("1st_ortho", times = length(RVConsensusOPLSModelCV$model$Model$lambda[,2]))),
                             "Blocks" = c(data_two_plots$Blocks,
                                          labels(collection)))

ggplot2::ggplot(data = data_two_plots,
                aes(x = factor(Type), y = Values, fill = factor(Type))) +
  ggplot2::geom_bar(stat = 'identity') + 
  ggplot2::xlab("Data blocks") +
  ggplot2::ylab("Weight") +
  ggplot2::ggtitle("Block contributions to each component")+
  facet_wrap(. ~ Blocks)+
  theme_graphs+
  ggplot2::scale_fill_discrete(name = "Component")+
  ggplot2::scale_fill_manual(values = c("#7F3C8D", "#11A579"))
```



## Block contributions predictive vs. orthogonal

```{r ggplot_data_pred_vs_ortho, class.source='fold-hide'}
ggplot2::ggplot(data = data.frame("Pred" = RVConsensusOPLSModelCV$model$Model$lambda[,1],
                                  "Ortho" = RVConsensusOPLSModelCV$model$Model$lambda[,2],
                                  "Labels" = labels(collection)),
                aes(x = Pred, y = Ortho, label = Labels, 
                    shape = Labels, colour = Labels)) +
  ggplot2::xlim(0, 0.5)+
  ggplot2::ylim(0, 0.5)+
  ggplot2::xlab("Predictive component") +
  ggplot2::ylab("Orthogonal component") +
  ggplot2::ggtitle("Block contributions predictive vs. orthogonal")+
  ggplot2::geom_point(size = 2.5) + 
  ggrepel::geom_text_repel(size = 4, show.legend = FALSE) + 
  theme_graphs+
  ggplot2::scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3"))
```

## Loading plots (one for each table)

```{r create_data_loadings}
loadings <- RVConsensusOPLSModelCV$model$Model$loadings
data_loads <- sapply(X = 1:length(collection),
                     FUN = function(X){
                       data.frame("Pred" = loadings[X,1][[1]],
                                  "Ortho" = loadings[X,2][[1]][[1]],
                                  "Labels" = labels(collection)[[X]])
                     })
data_loads <- as.data.frame(data_loads)
```

```{r ggplot_data_loadings, class.source='fold-hide'}
ggplot2::ggplot(data = data_loads) +
  ggplot2::geom_point(data = as.data.frame(data_loads$V1),
                      aes(x = Pred, y = Ortho, colour = Labels), 
                      size = 2.5, alpha = 0.5) + 
  ggplot2::geom_point(data = as.data.frame(data_loads$V2),
                      aes(x = Pred, y = Ortho, colour = Labels),
                      size = 2.5, alpha = 0.5) +
  ggplot2::geom_point(data = as.data.frame(data_loads$V3),
                      aes(x = Pred, y = Ortho, colour = Labels),
                      size = 2.5, alpha = 0.5) +
  ggplot2::xlim(-4, 4)+
  ggplot2::ylim(-3, 3)+
  ggplot2::xlab("Predictive component") +
  ggplot2::ylab("Orthogonal component") +
  ggplot2::ggtitle("ConsensusOPLS Loading plot")+
  theme_graphs+
  ggplot2::scale_color_manual(values = c("#1B9E77", "#D95F02", "#7570B3"))
```





# Permutations

```{r run_permutations, warning=FALSE}
# data = collection; Y = demo_data$Y; nbruns = 1000; PredLVs = LVsPred; maxOrtholvs = LVsOrtho;
PermRes <- RVConsensusOPLSPerm(data = collection,
                               Y = demo_data$Y,
                               nbruns = 1000,
                               PredLVs = LVsPred,
                               maxOrtholvs = LVsOrtho)
```

The execution time for this part of the method takes `r PermRes$execution_time`.

```{r plot1_perm}
PermRes$Plots$R2val
```

```{r plot2_perm}
PermRes$Plots$Q2val
```

```{r plot3_perm}
PermRes$Plots$R2_and_Q2
```

```{r plot4_perm}
PermRes$Plots$R2val_hist
```

```{r plot5_perm}
PermRes$Plots$Q2val_hist
```

```{r plot6_perm}
PermRes$Plots$R2_and_Q2_hist
```















# Reproducibility

Here is the output of `sessionInfo()` on the system on which this document was compiled:

```{r reproducibility}
sessionInfo()
```

# References
